---
title: "Capstone Project - Draft 1"
author: "John Morse"
bibliography: capstone.bib
csl: american-political-science-association.csl
link-citations: true
abstract: >
 ABSTRACT: Literacy is a foundational skill strongly associated with long-term academic and postsecondary outcomes. In Washington State, recent reductions in early literacy funding and stagnant college enrollment rates have renewed attention on how literacy achievement is measured and how it relates to postsecondary success. This study examines the relationship between secondary school literacy achievement, as measured by English Language Arts (ELA) scores on the Smarter Balanced Assessment Consortium (SBAC) exams, and enrollment in four-year colleges at the school district level. Using publicly available data from the Washington State Open Data Portal and the Education Research & Data Center, district-level SBAC ELA performance and college enrollment rates are analyzed through regression models. By assessing whether SBAC ELA outcomes are predictive of postsecondary enrollment, this study evaluates the relevance of existing standardized assessments for understanding college readiness and informs how state and district education agencies might better target literacy interventions to support increased college enrollment.

papersize: letter

margin-left: 1in
margin-right: 1in
margin-top: 1in
margin-bottom: 1in
cap-location: top

format: 
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    linestretch: 2
    pagestyle: plain
    
execute:
  echo: false
  warning: false
  message: false
---

\clearpage

\raggedright

\setlength{\parindent}{20pt}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(glue)
library(knitr)
library(kableExtra)
library(patchwork)
library(plm)
library(ggthemes)
library(huxtable)
library(jtools)
```

```{r}
#| label: source_datasets
#| include: false

##
# Source the R file that builds the necessary data sets.
source("build-dataset.R")
# Source the R code to build county datasets
source("build-county-dataset.R")
```

# Introduction

Literacy provides the base skillset of all student education, both through a student’s school years as well as beyond. In 2012, the Annie E. Casey Foundation published research on the importance of early literacy skills and poverty rates. Students who have not met expected standards in literacy by 3^rd^ grade will quickly fall behind, risking even their likelihood of graduation from high school [@Hernandez2011]. Yet programs designed to build literacy skills in childhood are often at risk.

In Washington state, budget cuts from the state Legislature in 2025 eliminated the Imagination Library program, an early learning program providing books for children from birth to age 5 [@Lotmore2025]. The Office of the Superintendent of Public Education (OSPI) has stepped in with discretionary funds for the current school year (2025-26), but the future of the program in Washington state remains unclear. A similar program, Prime Time Family Reading, encouraging reluctant readers and families to improve reading skills, lost 85% of funding through federal and state budget cuts [@Radil2026].

Literacy itself has a broad definition and can include word decoding, reading speed, text comprehension, reading strategies, and even motivation [@Jackson2005; @Talwar2022]. At its core, though, are the skills of reading, reading comprehension, and writing. At the secondary school level, Washington state uses standardized tests provided by the Smarter Balanced Assessment Consortium (SBAC) to track student achievement in English Language Arts (ELA) and Mathematics. Tests can be administered as many as eight times in a student’s secondary schooling, from grades 3 to 11.

These test scores provide a thread throughout secondary schooling where we can track the rate at which students are achieving expected literacy goals. At the same time, they measure the success of teaching and learning programs implemented within the schools and school districts. In fact, standardized tests are a direct result of the No Child Left Behind (NCLB) and Every Student Succeeds Act (ESSA) implemented by the Bush and Obama administrations in 2002 and 2015 respectively.

Although student success can be represented in many ways, including steady employment or military enrollment, 4-year college enrollment is a well-measured indicator of post-secondary outcomes. According to the Education Research & Data Center (ERDC) for Washington State, the rate of enrollment in 4-year colleges has remained nearly flat over the past 10 years, while enrollment in 2-year colleges and technical schools has decreased by four percentage points. At the same time, ERDC reports rates of enrollment in college-prep English courses has dropped from 9% to 6% for 4-year college destined students and from 36% to 15% for students expecting to enter 2-year colleges or technical schools [@ERDC2026].

With the decrease in enrollment, the Washington Student Achievement Council (WSAC) has set a goal that "70 percent of adults ages 25 to 44 should earn a post-secondary credential" [@WSAC2026Plan]. In the 10 years ending in 2024, though, WSAC measured an 18% decrease in post-secondary enrollment [@WSAC2026Dashboard].

This paper seeks to understand the relationship between student achievement in literacy at the school district level, and outcomes of that achievement in post-secondary success. Specifically, I will be assessing the relationship between ELA scores on SBAC testing and enrollment in 4-year colleges. With literacy capability in secondary school strongly associated with college readiness, tracking literacy achievement through SBAC scores may be a strong indication of college enrollment.

# Background

Research has long established a close relationship between literacy and academic success. “An ability to read, and to learn from reading, is a fundamental academic skill and its importance to scholastic success in any area of study at all levels of education is widely recognized.” [@Cox2003] Most recent discussions emphasize the link between early literacy and secondary education, notably high school graduation. Third-grade is frequently identified as a key point for reading skills in determining not just success, but graduation from high school. As much as 16% of children without proficient reading skills after third grade fail to graduate from high school on time, a statistic exacerbated further by a student’s socioeconomic status [@Hernandez2011]. Comprehensive literacy programs prior to third grade, however, show substantial increases in reading scores on standardized tests used by states for evaluating outcomes [@Westall2023].

The link continues between secondary school literacy skills and post-secondary outcomes. “\[Adults\] with a history of reading disability \[are\] less likely to pursue further education and more likely to spend a significant time in unemployment” [@James2024]. More directly, students with weak reading comprehension skills show lower scores on SATs, test results often used to determine college enrollment. Interestingly, reading comprehension weaknesses have a more pronounced impact on math and science test results than on ELA results [@James2024; @Westall2023].

The topic is hardly unique to the United States or the English language. A 2014 study based in Spain identified student literacy, again through standardized testing and found that reading skills and reading comprehension capability measured in earlier grades affects outcomes of later grades. Students in late primary (fifth and sixth grades) and early secondary (i.e. high school) grades showed natural improvement in reading comprehension with age, but those with reading difficulties in earlier grades continued to show weaker results in later grades [@Fidalgo2014]. Similarly, research in Finland found that reading and spelling capability assessed at the end of primary school (9^th^ grade) predicted secondary school choices, tracking students into either general education (a college-bound track) or vocational education [@Savolainen2008]. The assessment of literacy skills in earlier grades (i.e. US secondary school) ultimately predicts college enrollment and post-secondary outcomes. As @Talwar2022 indicate, “many of the factors that contribute to academic reading performance are also predictive of general academic success in college.”

Despite the clear impact of literacy skills on student outcomes both in high school and beyond, there is evidence that colleges may have set aside the need for literacy standards for student acceptance. A 2010 comparison of the reading comprehension of first-year university students at the University of Prince Edward Island to the broader population of Canada, found no difference in skills, indicating literacy skills were not deemed relevant to college enrollment [@Graham2010]. Five years earlier, a study of nearly 200 students at the University of Iowa found no strong relationship between literacy skills and academic success [@Jackson2005]. The question is, then, how important are reading and reading comprehension skills to college enrollment and college success?

The University of Iowa student population was restricted to second and third-year students in a single educational psychology class. The authors themselves describe them as “mature, successful students” [@Jackson2005], thus nullifying the analysis as to whether their literacy skills lead to academic success. Similarly, the research in Canada was limited to just over 220 students attending a single university in Eastern Canada. Limited samples prevent a definitive answer on the importance of literacy of secondary school students and their post-secondary opportunities.

This brings into question, are reading comprehension scores on standardized tests valid for assessing whether students are ready for college? Instead of depending on this assessment of reading skills to measure readiness, can we instead bring them up to a level of readiness to ensure success? Students entering college without the requisite literacy skills are often required to enroll in developmental or remedial reading courses as part of their first term. Completing such courses proves beneficial, as those students, previously unprepared, went on to complete college at the same rate as their peers [@Cox2003]. If a Canadian university waives the need for literacy skills to accept students, can we simply integrate literacy learning into college?

Conversely, remedial courses introduced at the secondary school level may have a detrimental effect on college enrollment. Based on standardized test scores, students in Florida required to enroll in remedial reading courses in high school are 5.4% less likely to enroll in either a 2-year or 4-year college program [@Ozek2025]. An unfortunate side effect of early assignment to a remedial course is that students may get “tracked” in remedial courses in general, resulting in fewer opportunities to succeed elsewhere in their education.

As indicated above, a common thread to the assessment of student literacy at all levels is the development and application of standardized tests of reading and reading comprehension. Though standardized testing is hardly new, attention to testing increased in the United States with the No Child Left Behind (NCLB) act in 2002 implemented by the Bush administration. NCLB was updated in 2015 as the Every Student Succeeds Act (ESSA) by the Obama administration. A key requirement of ESSA (and NCLB before it) was the implementation of standardized testing for reading and math in grades 3 through 8, and at least once in high school [@us_congress_essa_2015].

Perhaps there is no stronger argument for the importance of literacy skills testing than in a 2022 article by Amani Talwar et. al. “In the current landscape of higher education, academic reading tasks are important components of college courses and performance over time.” [@Talwar2022]. This research team chose to develop their own series of tests to apply to test subjects. Outcomes of their research indicate vocabulary and reading comprehension, commonly evaluated in standardized tests, as important skills for predicting early college success.

Research relating standardized test results to college enrollment has primarily emphasized college acceptance tests like the ACT and SAT. Literacy in third grade correlates with ELA standardized tests results (e.g. ACT) in middle grades, which, in-turn, relate to enrollment in honors and dual-credit courses [@Hein2013]. Those advanced courses, e.g. Advanced Placement (AP) and International Baccalaureate (IB) are subsequently correlated with college enrollment. But with universities waving the requirement for college entrance exams in the wake of the pandemic [@Schwartz2021; @BauerWolf2020], these tests may no longer have relevance in assessing college readiness for students. So how do we judge a student’s college readiness? Although some research has indicated high school GPAs may provide a strong predictor of college success [@Kurlaender2019], an overall GPA may hide weaknesses in a specific subject. GPAs in college-preparatory courses may eliminate some students from eligibility.

SAT and ACT scores are specifically designed for college entrance assessment, and will have, by design, a skewed student population already seeking college. Similarly, AP scores and similar dual-credit courses occur most often in the last two years of secondary school, also attracting students already targeting college. SAT, ACT, and AP testing students are, by nature, preparing for these standardized tests. In fact, AP courses are designed specifically for preparing students to take AP courses. The tests themselves are designed specifically as an assessment of college readiness [@Talwar2022]. Some evidence indicates a high school GPA is a stronger predictor of college readiness than college entrance tests like the SAT and ACT [@Geiser2007]. SBAC testing, on the other hand, has been shown to be as relevant as the SAT in predicting college success [@Kurlaender2019]. Although the same report indicated GPA may be a better predictor of an individual student’s success in college, a GPA can not be used to evaluate the success of programs implemented across a school district.

If we are to increase college enrollment to 70% of the adult population, as WSAC has targeted, we need to seek opportunities beyond the self-selected college bound population. In Washington state, the SBAC tests are not subject to the focus of test-taking strategies like the SAT or ACT might be. Most students are given the SBAC simply as part of the expected school year. They do not study or prepare for the test outside of "teaching to the test" strategies that may occur as part of the standard school curriculum. However, those strategies would be applied equally across all students. Compare this to an AP course, for example, where a select number of students are choosing to take a course centered on the subject matter of the AP exam.

The NCLB (and subsequently ESSA) act specifically requires standardized testing for the purpose of identifying differences in the quality of education provided by different schools and school districts [@Wiliam2010], rewarding those that do well. With that in mind, the SBAC ELA tests are a direct evaluation of the literacy programs in schools. With literacy identified as a primary skill for college success, and SBAC tests identified as a primary means of measuring literacy programs in secondary schools, we can expect a relationship between the SBAC ELA outcomes in a school district and the rate of enrollment in college of students in that district.

This paper examines the results of standardized tests given to secondary school students in Washington state on an annual basis. By demonstrating the relevance of existing annualized testing on college enrollment, state and district educational organizations can use the information to more directly understand where gaps exist in literacy achievement and apply critical funding as appropriate. Increased enrollment in secondary education can only occur with a clear picture of where schools and school districts are or are not driving enrollment.

# Data and Methodology

In this project, I am seeking to understand whether the SBAC ELA scores predict college enrollment rates at the school district level. Based on this hypothesis, Washington state could track progress within school district toward meeting state goals of post-secondary enrollment.

## Data

Washington student achievement on SBAC test results, along with enrollment in college, are reported as ratios through the Washington State Open Data Portal (<https://data.wa.gov>) and through online dashboards provided by ERDC. Through regression analysis, I will analyze SBAC test results in respect to college enrollment rates for school districts throughout the state. Additionally, I will examine the effect of Math scores on those same enrollment rates to determine if there is a significant difference in influence.

I will be working with overall rates of enrollment and rates of meeting test standards at the school and school district level. This introduces a potential risk of an ecological fallacy, as there may be several hidden factors that exist at the school or student level. @Wiliam2010 previously determined that socioeconomic factors may, in fact, influence differences between schools more than the actual differences in education. In an effort to control for these risks, I will introduce entity-fixed effects into the regression, along with additional test results for low-income identified students.  Additionally, the time period considered, 2014-15 to 2023-24 school years includes the COVID-19 pandemic, introducing the need for time-fixed effects in the analysis.

SBAC scores from a single year (e.g. 11^th^ grade) may not be wholly representative of student achievement and readiness. To validate the robustness of the relationship between SBAC achievement scores and post-secondary enrollment, I intend to analyze multiple perspectives of standardized test scores:

-   Most recent SBAC scores (last score reported in a school district).

-   Last two SBAC scores to smooth single test variations.

-   A simple aggregation of SBAC scores across all grades in a school district.

-   Weighted scores across all grades emphasizing later grades over earlier.

In all cases, SBAC test scores are reported as the rate at which students met or exceeded expected standards for their grade.

This analysis attempts to control for various socioeconomic and demographic charateristics including employment, median household income, and health insurance. Consideration of variables outside schooling and academics has multiple aspects of importance. As @Geiser2007 indicates, many universities give extra consideration towards applicants from disadvantaged backgrounds or schools. Additionally, analysis at the school district level may introduce ecological fallacies due to aggregated results, and inclusion of socioeconomic controls can account for OVB. Finally, standardized test results can be correlated with background factors, reflecting student environment outside of school.

Variables that have been left out of the analysis include x, y, and z because blah blah &&&.

## Methodology

I will be tracking the rates at which students are:

-   Entering college (2-year or 4-year)

-   Control factor: Participating in remedial courses in the first year. Theoretically this is an indication that the student is NOT ready for college.

This will be measured based on influencing variables such as:

-   Rate at which students meet standards for ELA testing.

-   Rate at which test results within a District change over time.

There are many factors that can influence a child's readiness for reading and writing throughout their schooling, including income levels, environment (e.g. urban vs. rural), gender, and ethnicity. For the purpose of this project, I will attempt to control for low-income status and gender, though recognize that additional factors may be hidden in the data. My goal is to assess how reading & writing ability, as measured through standardized testing, predict enrollment in 4-year colleges. Math scores will also be considered as a contributing factor to enrollment rates.

My primary assessment of literacy capability will be based on the percent of students meeting expected standards for the ELA SBAC test. This test is administered 8 times throughout a student's secondary school education. The OSPI data includes the total number of students testing, the total number meeting standards, and the percent meeting standards. An SBAC score is a value of 1 to 4, with values of 3 and 4 indicating the student met expected standards.

The data includes multiple school years and grade levels. However, as this research question does not examine changes across time, district level results are aggregated into a single indicator for the district.

The OSPI data includes scores for individual schools and for each grade, indicating the percent of students meeting expectations within the grade for that school year. Aggregates are also provided at the All Grades level and at the school district level. I have chosen to create an aggregate score for each school year and for each school district based on the percentages provided for all grades and all schools within the district. This is handled using a Logit transformation of weighted scores. I am weighting the grades so that later grades have more influence on the overall literacy score, indicating more influence on college enrollment and persistence.

$$ \text{logit}_{g} = \ln\!\left(\frac{p_{g}}{1 - p_{g}}\right) $$

$$ w_{g} = \text{grade\_weight}_{g} \cdot n_{g} $$

$$ \text{WeightedLogit}_{d,y} = \frac{ \sum_{g} \left( \text{logit}_{g} \cdot w_{g} \right) }{ \sum_{g} w_{g} } $$

Above are formulas rendered in Latex

Enrollment rates for each year are aggregated at the School District level using a Logit_average technique. I did not have access to the underlying numbers for students enrolled in each year. Thus, the ratios provided in my source data needed to be averaged without weighting. The Logit-average technique is appropriate for averaging ratios.

Formulas:

Logit-average formula converts each rate of enrollment into a logit value using the natural logarithm. Values across years are averaged.

$$
\text{LogitAvg} 
= \frac{1}{T} \sum_{t=1}^{T} \operatorname{logit}(r_t)
$$

Once the values are averaged, the value is converted back into a standard ratio value.

$$
R = \operatorname{logistic}(\text{LogitAvg})
= \frac{1}{1 + e^{-\text{LogitAvg}}}
$$

## Fixed Effects Regression

Pooling ignores:

-   Differences across districts

-   time-based influence across districts (e.g. Covid pandemic)

\<NOTE: REWRITE THIS\> Inclusion of ela_li_met in the model helps separate:

-   “districts with high overall ELA proficiency” from

-   “districts where low‑income students specifically perform well” .

Controls for

-   differences in low-income student performance across districts

-   changes in low-income performance over time

-   possibility of subgroup perofmrance confounding ela_met

It does not control for: • the share of low‑income students in the district • poverty rates • socioeconomic composition

-   Look for % low-income or % FRPL

# Analysis

As an initial analysis of how ELA SBAC assessment results relate to college enrollment, the following chart plots each individual school district for each year and applies a non-linear regression line.

### Scatter Plots

```{r}
#| label: scatter_code
#| include: false
p1 <- ggplot(data = dfCapAllGrades, 
             aes(x = ela_met, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = "ELA expectations captured across all grades within a School District.",
    x = NULL,
    y = "% Enrolling in College"
  )

p2 <- ggplot(data = dfCapLastGrade, 
             aes(x = ela_met, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = "ELA expectations captured for Last Grade Tested within a School District.",
    x = NULL,
    y = "% Enrolling in College"
  )

p3 <- ggplot(data = dfCapLast2Grades, 
             aes(x = ela_met, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = "ELA expectations captured for Last 2 Grades Tested within a School District.",
    x = NULL,
    y = "% Enrolling in College"
  )

p4 <- ggplot(data = dfCapWeighted, 
             aes(x = ela_score, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = "ELA expectations weighted across all grades tested within a School District.",
    x = "% Meeting SBAC ELA Expectations",
    y = "% Enrolling in College"
  )
```

```{r}
#| label: scatter_plot
#| echo: false
#| fig-cap: College Enrollment and ELA Assessment
#| fig-height: 9.5
#| fig-width: 7

combined_plot <- p1 / p2 / p3 / p4
#combined_plot <- combined_plot +
#  plot_annotation(
#    title = "Plot of College Enrollment and Students Meeting ELA Expectations"
#  )
combined_plot
```

\clearpage

### Density Plots

Next let's look at a series of density plots, one for each school year.

```{r}
#| label: density_code
#| include: false
# Focus on just one data set: weighted assessments

# Create a data set with median values for each subject
median_enroll <- dfCapWeighted %>%
  group_by(year) %>%
  summarize(xbar = median(enrollment_all, na.rm=TRUE))
#median_per_enroll$y <- 0.3  # Used to position label

# facet_grid by year and by test_subject
p1 <- ggplot(data = dfCapWeighted, 
            mapping = aes(x = enrollment_all)) +
  geom_density(alpha = 1, fill = "darkblue",
                 mapping = aes(y = after_stat(scaled))) + 
  geom_vline(data = median_enroll, aes(xintercept = xbar),
             color = "white", linewidth = 0.5) +
  geom_text(data = median_enroll,
            aes(x = xbar - 0.025, y = 0.2,
                label = str_c(format(xbar * 100, digits = 1), "%")),
                size = 3, hjust = 1, color = "white") +
  scale_x_continuous(breaks = c(.25, .50, .75), 
                     labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  labs(subtitle = "Enrollment",
       x = NULL, y = NULL
       #caption = "Enrollment from 2015 to 2023."
  ) +
  facet_grid(year ~ ., switch = "y")

median_assess <- dfCapWeighted %>%
  group_by(year) %>%
  summarize(xbar = median(ela_score, na.rm=TRUE))

p2 <- ggplot(data = dfCapWeighted,
             mapping = aes(x = ela_score)) +
  geom_density(alpha = 1, fill = "darkred",
                 mapping = aes(y = after_stat(scaled))) + 
  geom_vline(data = median_assess, aes(xintercept = xbar),
             color = "white", linewidth = 0.5) +
  geom_text(data = median_assess,
            aes(x = xbar - 0.02, y = 0.2,
                label = str_c(format(xbar * 100, digits = 1), "%")),
                size = 3, hjust = 1, color = "white") +
  scale_x_continuous(breaks = c(.25, .50, .75), 
                     labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  labs(subtitle = "Weighted Assessment",
       x = NULL, y = NULL
       #caption = "Enrollment from 2015 to 2023."
  ) +
  facet_grid(year ~ ., switch = "y")

median_assess2 <- dfCapAllGrades %>%
  group_by(year) %>%
  summarize(xbar = median(ela_met, na.rm=TRUE))

p3 <- ggplot(data = dfCapAllGrades,
             mapping = aes(x = ela_met)) +
  geom_density(alpha = 1, fill = "indianred",
                 mapping = aes(y = after_stat(scaled))) + 
  geom_vline(data = median_assess2, aes(xintercept = xbar),
             color = "white", linewidth = 0.5) +
  geom_text(data = median_assess2,
            aes(x = xbar - 0.02, y = 0.2,
                label = str_c(format(xbar * 100, digits = 1), "%")),
                size = 3, hjust = 1, color = "white") +
  scale_x_continuous(breaks = c(.25, .50, .75), 
                     labels = scales::percent) +
  scale_y_continuous(labels = NULL) +
  labs(subtitle = "All Grades Assessment",
       x = NULL, y = NULL
       #caption = "Enrollment from 2015 to 2023."
  ) +
  facet_grid(year ~ ., switch = "y")
 
```

```{r}
#| label: density_plot
#| echo: false
#| fig-cap: Distribution of Assessment and Enrollment by Year
#| fig-height: 6
#| fig-width: 7

combined_plot <- p3 | p2 | p1
combined_plot
```

\clearpage

### Boxplots

Next are boxplots by year, one for each assessment type.

```{r}
#| label: boxplot_code
#| include: false

# All grades
p1 <- ggplot(data = dfCapAllGrades,
             aes(x = factor(year), y = ela_met)) +
  geom_boxplot(color = "darkred") +
  theme(plot.subtitle = element_text(face="bold")) +
  labs(
    subtitle = "All grades in school district",
    x = NULL,
    y = NULL
  )

# Most recent Grade
p2 <- ggplot(data = dfCapLastGrade,
             mapping = aes(x = factor(year), y = ela_met)) +
  geom_boxplot(color = "darkred") +
  theme(plot.subtitle = element_text(face="bold")) +
  labs(
    subtitle = "Last grade tested in school district",
    x = NULL,
    y = NULL
  )

# Last 2 grades
p3 <- ggplot(data = dfCapLast2Grades,
             mapping = aes(x = factor(year), y = ela_met)) +
  geom_boxplot(color = "darkred") +
  theme(plot.subtitle = element_text(face="bold")) +
  labs(
    subtitle = "Last grade tested in school district",
    x = NULL,
    y = NULL
  )

# Weighted across grades
p4 <- ggplot(data = dfCapWeighted,
             mapping = aes(x = factor(year), y = ela_score)) +
  geom_boxplot(color = "darkred") +
  theme(plot.subtitle = element_text(face="bold")) +
  labs(
    subtitle = "Last grade tested in school district",
    x = "Year",
    y = NULL
  )

# Enrollment
p5 <- ggplot(data = dfCapAllGrades,
             mapping = aes(x = factor(year), y = enrollment_all)) +
  geom_boxplot(color = "darkblue") +
  labs(
    #title = "Distribution of SBAC ELA Assessment by Year",
    subtitle = "College Enrollment",
    x = "Year",
    y = "Percent Enrolled"
  )

```

```{r}
#| label: boxplots
#| echo: false
#| fig-cap: Distribution of Assessment and Enrollment by Year
#| fig-height: 8
#| fig-width: 7

combined_plot <- p1 / p2 / p3 / p4 #/ p5
combined_plot
```

### Heatmap

Heatmap by district with randomly selected list of 30 school districts.

```{r}
#| label: heatmap_code
#| include: false

# Randomly select 30 school districts
set.seed(456)
sampled_districts <- dfCapAllGrades %>%
  distinct(district_code) %>%
  slice_sample(n = 30) %>%
  pull(district_code)

df1 <- dfCapAllGrades %>%
  filter(district_code %in% sampled_districts)

# Generate a heatmap plot
p1 <- ggplot(data = df1,
             aes(x = factor(year), 
                 y = district_name, 
                 fill = ela_met)
             ) +
  geom_tile() +
  # Put x-axis at top of graph
  scale_x_discrete(position = "top") + 
  # sort the school districts
  scale_y_discrete(limits = rev(sort(unique(df1$district_name)))) +
  scale_fill_gradient2(midpoint=0.5,
                       low = "darkred", high = "darkblue",
                       name = "Percent Met Standards", 
                       label = scales::percent,
                       breaks = c(0.2, 0.5, 0.8)) +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.box.margin = margin(0, 0, 5, 0),
        legend.text = element_text(size = rel(0.8)),
        legend.title = element_text(size = rel(0.9)),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size=rel(0.8)),
        axis.text.y = element_text(size=rel(0.9)),
        plot.subtitle = element_text(size=rel(1.0), face="bold")
        ) +
  labs(title = NULL,
       subtitle = "For random selection of 30 school districts",
       caption = "Darker reds = lower rates of students meeting standards.",
       )

```

```{r}
#| label: heatmap_chart
#| echo: false
#| fig-cap: SBAC ELA Trends by Year
#| fig-height: 8
#| fig-width: 7

p1
```

Primary pattern here is that those districts with higher scores tend to always have higher scores, while those with lower scores tend to always have lower scores. Although some districts show inconsistency, and in many cases we see a sudden drop in the last two years, after COVID. Overall, though, this implies a consistency within the district from year to year, potentially supporting the targeting of school districts that need further help to improve SBAC ELA results.

\clearpage

### Mapping

Create a map of the assessment results, highlighting counties by percent of students meeting standards.

```{r}
#| label: mapping_code
#| include: false

# Get Map data for Washington state counties.
# This is part of ggplot2.
wa_counties <- map_data("county", "washington")

# Create a data set with county map data and the graduation rates
dfMapAllGrades <- wa_counties %>%
  # Get ALL Grades data from build-county-dataset.R
  # NOTE: County names must be in all lower case.
  left_join(df_assess_allgrades_cty, by = c("subregion" = "county"))

p1 <- ggplot(data = dfMapAllGrades,
             mapping = aes(x = long, y = lat, 
                           group = group, fill = ela_met)) +
  geom_polygon(color = "black") +
  scale_fill_gradient2(midpoint=0.5,
                       low = "darkred", high = "darkblue",
                       name = "Percent Met Standards", 
                       label = scales::percent) +
  facet_wrap(~school_year, ncol = 2) +
  theme_map() + 
  theme(legend.position = "top",
        plot.title = element_text(size=rel(1.5), face="bold")) +
  labs(title = "SBAC ELA Assessment Rates in Washington by County",
       subtitle = "",
       caption = "Darker red = lower rates of students meeting standards.\nGray color indicates data not available for that county.")
```

```{r}
#| label: mapping_plot
#| echo: false
#| fig-cap: Mapping of Assessment Rates by County
#| fig-height: 9.5
#| fig-width: 7

p1
```

\clearpage

### Regression

Let's begin doing some regression.

\begin{center}
Table 1: Pooled Regression of Enrollment by SBAC Assessment
\end{center}

```{r}
#| label: pooled_regression
#| echo: false
# Create a pdata.frame with entity and time indexes
pdata_allgrades <- pdata.frame(dfCapAllGrades, 
                               index = c("district_code", "year"))
pdata_lastgrade <- pdata.frame(dfCapLastGrade,
                               index = c("district_code", "year"))
pdata_last2grades <- pdata.frame(dfCapLast2Grades,
                                 index = c("district_code", "year"))
pdata_weighted <- pdata.frame(dfCapWeighted, 
                              index = c("district_code", "year"))

# Start with pooled regression for each model
pool_allgrades <- plm(enrollment_all ~ ela_met,
                      data = pdata_allgrades,
                      model = "pooling")
pool_lastgrade <- plm(enrollment_all ~ ela_met,
                      data = pdata_lastgrade,
                      model = "pooling")
pool_last2grades <- plm(enrollment_all ~ ela_met,
                        data = pdata_last2grades,
                        model = "pooling")
pool_weighted <- plm(enrollment_all ~ ela_score,
                     data = pdata_weighted,
                     model = "pooling")

# Build a table with the output from export_summs
tbl <- export_summs(pool_allgrades, pool_lastgrade,
                    pool_last2grades, pool_weighted,
                    model.names = c("All Grades", "Last Grade",
                                    "Last 2 Grades", "Weighted"),
                    robust = "HC1"
                    )
# Format the table
tbl <- set_font_size(tbl, 8)
# show the table
tbl

```

# Discussion

Discuss the results

# Conclusion

Summarize and wrap up.

\clearpage

# References
