---
title: "Capstone Project - Draft 1"
author: "John Morse"
bibliography: capstone.bib
csl: american-political-science-association.csl
link-citations: true
abstract: >
 ABSTRACT: Literacy is a foundational skill strongly associated with long-term academic and postsecondary outcomes. In Washington State, recent reductions in early literacy funding and stagnant college enrollment rates have renewed attention on how literacy achievement is measured and how it relates to postsecondary success. This study examines the relationship between secondary school literacy achievement, as measured by English Language Arts (ELA) scores on the Smarter Balanced Assessment Consortium (SBAC) exams, and enrollment in four-year colleges at the school district level. Using publicly available data from the Washington State Open Data Portal and the Education Research & Data Center, district-level SBAC ELA performance and college enrollment rates are analyzed through regression models. By assessing whether SBAC ELA outcomes are predictive of postsecondary enrollment, this study evaluates the relevance of existing standardized assessments for understanding college readiness and informs how state and district education agencies might better target literacy interventions to support increased college enrollment.

papersize: letter

margin-left: 1in
margin-right: 1in
margin-top: 1in
margin-bottom: 1in
cap-location: top

format: 
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    linestretch: 2
    pagestyle: plain
    include-in-header:
      text: |
        \usepackage{float}

execute:
  echo: false
  warning: false
  message: false
---

\clearpage

\raggedright

\setlength{\parindent}{20pt}

```{r}
#| label: libraries
#| include: false
library(tidyverse)
library(glue)
library(plm)
library(huxtable)
library(jtools)
library(gt)
library(sandwich)
library(lmtest)
```

```{r}
#| label: source_datasets
#| include: false

##
# Source the R file that builds the necessary data sets.
source("build-dataset.R")
```

# Introduction

Literacy provides the base skill set of all student education, both through a student’s school years as well as beyond. In 2012, the Annie E. Casey Foundation published research on the importance of early literacy skills and poverty rates. Students who have not met expected standards in literacy by 3^rd^ grade will quickly fall behind, risking even their likelihood of graduation from high school [@Hernandez2011]. Yet programs designed to build literacy skills in childhood are often at risk.

In Washington state, budget cuts from the state Legislature in 2025 eliminated the Imagination Library program, an early learning program providing books for children from birth to age 5 [@Lotmore2025]. The Office of the Superintendent of Public Education (OSPI) has stepped in with discretionary funds for the current school year (2025-26), but the future of the program in Washington state remains unclear. A similar program, Prime Time Family Reading, encouraging reluctant readers and families to improve reading skills, lost 85% of funding through federal and state budget cuts [@Radil2026].

Literacy itself has a broad definition and can include word decoding, reading speed, text comprehension, reading strategies, and even motivation [@Jackson2005; @Talwar2022]. At its core, though, are the skills of reading, reading comprehension, and writing. At the secondary school level, Washington state uses standardized tests provided by the Smarter Balanced Assessment Consortium (SBAC) to track student achievement in English Language Arts (ELA) and Mathematics. Tests can be administered as many as eight times in a student’s secondary schooling, from grades 3 to 11.

These test scores provide a thread throughout secondary schooling where we can track the rate at which students are achieving expected literacy goals. At the same time, they measure the success of teaching and learning programs implemented within the schools and school districts. In fact, standardized tests are a direct result of the No Child Left Behind (NCLB) and Every Student Succeeds Act (ESSA) implemented by the Bush and Obama administrations in 2002 and 2015 respectively.

Although student success can be represented in many ways, including steady employment or military enrollment, 4-year college enrollment is a well-measured indicator of post-secondary outcomes. According to the Education Research & Data Center (ERDC) for Washington State, the rate of enrollment in 4-year colleges has remained nearly flat over the past 10 years, while enrollment in 2-year colleges and technical schools has decreased by four percentage points. At the same time, ERDC reports rates of enrollment in college-prep English courses has dropped from 9% to 6% for 4-year college destined students and from 36% to 15% for students expecting to enter 2-year colleges or technical schools [@ERDC2026].

With the decrease in enrollment, the Washington Student Achievement Council (WSAC) has set a goal that "70 percent of adults ages 25 to 44 should earn a post-secondary credential" [@WSAC2026Plan]. In the 10 years ending in 2024, though, WSAC measured an 18% decrease in post-secondary enrollment [@WSAC2026Dashboard].

This paper seeks to understand the relationship between student achievement in literacy at the school district level, and outcomes of that achievement in post-secondary success. Specifically, I will be assessing the relationship between ELA scores on SBAC testing and enrollment in 4-year colleges. With literacy capability in secondary school strongly associated with college readiness, tracking literacy achievement through SBAC scores may be a strong indication of college enrollment.

# Background

Research has long established a close relationship between literacy and academic success. “An ability to read, and to learn from reading, is a fundamental academic skill and its importance to scholastic success in any area of study at all levels of education is widely recognized.” [@Cox2003] Most recent discussions emphasize the link between early literacy and secondary education, notably high school graduation. Third-grade is frequently identified as a key point for reading skills in determining not just success, but graduation from high school. As many as 16% of children without proficient reading skills after third grade fail to graduate from high school on time, a statistic exacerbated further by a student’s socioeconomic status [@Hernandez2011]. Comprehensive literacy programs prior to third grade, however, show substantial increases in reading scores on standardized tests used by states for evaluating outcomes [@Westall2023].

The link continues between secondary school literacy skills and post-secondary outcomes. “\[Adults\] with a history of reading disability \[are\] less likely to pursue further education and more likely to spend a significant time in unemployment” [@James2024]. More directly, students with weak reading comprehension skills show lower scores on SATs, test results often used to determine college enrollment. Interestingly, reading comprehension weaknesses have a more pronounced impact on math and science test results than on ELA results [@James2024; @Westall2023].

The topic is hardly unique to the United States or the English language. A 2014 study based in Spain identified student literacy, again through standardized testing and found that reading skills and reading comprehension capability measured in earlier grades affects outcomes of later grades. Students in late primary (fifth and sixth grades) and early secondary (i.e. high school) grades showed natural improvement in reading comprehension with age, but those with reading difficulties in earlier grades continued to show weaker results in later grades [@Fidalgo2014]. Similarly, research in Finland found that reading and spelling capability assessed at the end of primary school (9^th^ grade) predicted secondary school choices, tracking students into either general education (a college-bound track) or vocational education [@Savolainen2008]. The assessment of literacy skills in earlier grades (i.e. US secondary school) ultimately predicts college enrollment and post-secondary outcomes. As @Talwar2022 indicate, “many of the factors that contribute to academic reading performance are also predictive of general academic success in college.”

Despite the clear impact of literacy skills on student outcomes both in high school and beyond, there is evidence that colleges may have set aside the need for literacy standards for student acceptance. A 2010 comparison of the reading comprehension of first-year university students at the University of Prince Edward Island to the broader population of Canada, found no difference in skills, indicating literacy skills were not deemed relevant to college enrollment [@Graham2010]. Five years earlier, a study of nearly 200 students at the University of Iowa found no strong relationship between literacy skills and academic success [@Jackson2005]. The question is, then, how important are reading and reading comprehension skills to college enrollment and college success?

The University of Iowa student population was restricted to second and third-year students in a single educational psychology class. The authors themselves describe them as “mature, successful students” [@Jackson2005], thus nullifying the analysis as to whether their literacy skills lead to academic success. Similarly, the research in Canada was limited to just over 220 students attending a single university in Eastern Canada. Limited samples prevent a definitive answer on the importance of literacy of secondary school students and their post-secondary opportunities.

This brings into question, are reading comprehension scores on standardized tests valid for assessing whether students are ready for college? Instead of depending on this assessment of reading skills to measure readiness, can we instead bring them up to a level of readiness to ensure success? Students entering college without the requisite literacy skills are often required to enroll in developmental or remedial reading courses as part of their first term. Completing such courses proves beneficial, as those students, previously unprepared, went on to complete college at the same rate as their peers [@Cox2003]. If a Canadian university waives the need for literacy skills to accept students, can we simply integrate literacy learning into college?

Conversely, remedial courses introduced at the secondary school level may have a detrimental effect on college enrollment. Based on standardized test scores, students in Florida required to enroll in remedial reading courses in high school are 5.4% less likely to enroll in either a 2-year or 4-year college program [@Ozek2025]. An unfortunate side effect of early assignment to a remedial course is that students may get “tracked” in remedial courses in general, resulting in fewer opportunities to succeed elsewhere in their education.

As indicated above, a common thread to the assessment of student literacy at all levels is the development and application of standardized tests of reading and reading comprehension. Though standardized testing is hardly new, attention to testing increased in the United States with the No Child Left Behind (NCLB) act in 2002 implemented by the Bush administration. NCLB was updated in 2015 as the Every Student Succeeds Act (ESSA) by the Obama administration. A key requirement of ESSA (and NCLB before it) was the implementation of standardized testing for reading and math in grades 3 through 8, and at least once in high school [@us_congress_essa_2015].

Perhaps there is no stronger argument for the importance of literacy skills testing than in a 2022 article by Amani Talwar et. al. “In the current landscape of higher education, academic reading tasks are important components of college courses and performance over time.” [@Talwar2022]. This research team chose to develop their own series of tests to apply to test subjects. Outcomes of their research indicate vocabulary and reading comprehension, commonly evaluated in standardized tests, as important skills for predicting early college success.

Research relating standardized test results to college enrollment has primarily emphasized college acceptance tests like the ACT and SAT. Literacy in third grade correlates with ELA standardized tests results (e.g. ACT) in middle grades, which, in-turn, relate to enrollment in honors and dual-credit courses [@Hein2013]. Those advanced courses, e.g. Advanced Placement (AP) and International Baccalaureate (IB) are subsequently correlated with college enrollment. But with universities waving the requirement for college entrance exams in the wake of the pandemic [@Schwartz2021; @BauerWolf2020], these tests may no longer have relevance in assessing college readiness for students. So how do we judge a student’s college readiness? Although some research has indicated high school GPAs may provide a strong predictor of college success [@Kurlaender2019], an overall GPA may hide weaknesses in a specific subject. Alterantively, focusing on GPAs for only college-preparatory courses may eliminate students from eligibility if they are not in a traditional college-pre track in school.

SAT and ACT scores are specifically designed for college entrance assessment, and will have, by design, a skewed student population already seeking college. Similarly, AP scores and similar dual-credit courses occur most often in the last two years of secondary school, also attracting students already targeting college. SAT, ACT, and AP testing students are, by nature, preparing for these standardized tests. In fact, AP courses are designed specifically for preparing students to take AP tests. The tests themselves are designed specifically as an assessment of college readiness [@Talwar2022]. Some evidence indicates a high school GPA is a stronger predictor of college readiness than college entrance tests like the SAT and ACT [@Geiser2007]. SBAC testing, on the other hand, has been shown to be as relevant as the SAT in predicting college success [@Kurlaender2019]. Although the same report indicated GPA may be a better predictor of an individual student’s success in college, a GPA can not be used to evaluate the success of programs implemented across a school district.

If we are to increase college enrollment to 70% of the adult population, as WSAC has targeted, we need to seek opportunities beyond the self-selected college bound population. In Washington state, the SBAC tests are not subject to the focus of test-taking strategies like the SAT or ACT might be. Most students are given the SBAC simply as part of the expected school year. Though many teachers many provide curriculum that can be described as "teaching to the test," this curriculum is applied equally across all students and will be integrated into standard school curriculum. Compare this to an AP course, for example, where a select number of students are choosing to take a course centered on the subject matter of the AP exam.

The NCLB (and subsequently ESSA) act specifically requires standardized testing for the purpose of identifying differences in the quality of education provided by different schools and school districts [@Wiliam2010], rewarding those that do well. With that in mind, the SBAC ELA tests are a direct evaluation of the literacy programs in schools. With literacy identified as a primary skill for college success, and SBAC tests identified as a primary means of measuring literacy programs in secondary schools, we can expect a relationship between the SBAC ELA outcomes in a school district and the rate of enrollment in college of students in that district.

This paper examines the results of standardized tests given to secondary school students in Washington state on an annual basis. By demonstrating the relevance of existing annualized testing on college enrollment, state and district educational organizations can use the information to more directly understand where gaps exist in literacy achievement and apply critical funding as appropriate. Increased enrollment in post-secondary education can only occur with a clear picture of where school districts are successfully encouraging that enrollment. Insight into the success of the programs and skills necessary to encourage enollment can best be acquired through monitoring of progress.

# Data and Methodology

In this project, I am seeking to understand whether the ELA scores from SBAC tests administered within a school district provide insight into expected 4-year college enrollment within those districts. If confirmed, Washington state school officials could turn toward standardized tests to track progress within school districts toward meeting state goals of post-secondary enrollment.

There are many factors that can influence a child's readiness for reading and writing throughout their schooling, including income levels, environment (e.g. urban vs. rural), gender, and ethnicity. For the purpose of this project, I will attempt to control for low-income status and gender, though recognize that additional factors may be hidden in the data. My goal is to assess how reading & writing ability, as measured through standardized testing, predict enrollment in 4-year colleges. Math scores will also be considered as a contributing factor to enrollment rates.

## Sample

Washington student achievement on SBAC test results, along with enrollment in college, are reported as ratios through the Washington State Open Data Portal (<https://data.wa.gov>) and through online dashboards provided by ERDC. Data for test results and college enrollment were downloaded from the Open Data Portal as multiple CSV files, loaded into R and combined to create core data sets. A total of eight files were downloaded for SBAC assessment results covering the school years 2014-15 to 2018-19 and the years 2021-22 to 2023-24. Data for the 2023-24 school year can be accessed here: <https://data.wa.gov/education/Report-Card-Assessment-Data-2023-24-School-Year/x73g-mrqp/about_data>. Note that SBAC Assessments were not administered during the 2019-20 or 2020-21 school years due to the COVID-19 pandemic.

College enrollment data was downloaded as a single file covering the years 2005 to 2025. All data for college enrollment can be accessed through this site: <https://data.wa.gov/education/High-School-Graduate-Outcomes-First-Year-Enrollmen/vk6s-am8z/about_data>. Enrollment and SBAC Assessment data were joined based on school-district and enrollment year with adjustments to assessment data as described in the next section. Additionally, school and school district reference data was obtained through the Office of the Superintendent of Public Instruction (OSPI) here: <https://eds.ospi.k12.wa.us/DirectoryEDS.aspx>.

Control variables were obtained through the United States Census Bureau. Data was downloaded for the years 2014 to 2024 as 5-year estimates based on the American Community Survey. Data is collected over a period of five years and averaged through weighted adjustments to obtain the most accurate estimate of, for example, population, employment, and education. Data was obtained at the school district level for Selected Economic Characteristics (<https://data.census.gov/table?q=DP03&g=040XX00US53>) and Educational Attainment (<https://data.census.gov/table/ACSST1Y2024.S1501?q=education+attainment&g=040XX00US53,53$9700000>).

## Predictor Variables

My primary assessment of literacy capability will be based on the ratio of students meeting expected standards for the ELA SBAC test, identified as `ela_met` in the data. This test is administered 8 times throughout a student's secondary school education. The OSPI data includes the total number of students testing, the total number meeting standards, and the percent meeting standards. An SBAC score is a value of 1 to 4, with values of 3 and 4 indicating the student met expected standards. A secondary predictor variable examined here is the ratio of students meeting Level 4 standards, identified as `ela_lvl4`. Therefore, the ratio of students at Level 4 are representative of the highest achieving students on the SBAC assessment.

The OSPI data includes scores for individual school districts and for each grade, indicating the percent of students meeting expectations within the grade for that school year. Analysis at the school district level may introduce ecological fallacies due to aggregated results. @Wiliam2010 previously determined that inclusion of socioeconomic factors may, in fact, influence differences between schools more than the actual differences in education. Two approaches are considered in this document to account for potential Omitted Variable Bias (OVB). The first consideration is through entity-fixed and time-fixed effects based on panel regression. This will attempt to account for unknown variables affecting differences across school districts through the state. Additionally, the time period considered, 2014-15 to 2023-24 school years includes the COVID-19 pandemic, introducing the need for time-fixed effects in the analysis.

A second approach will be the introduction of several socioeconomic and demographic control variables as gathered from Census data. These include employment rates, median income, health insurance, and parent education. Consideration of variables outside schooling and academics has multiple aspects of importance. As @Geiser2007 indicates, many universities give extra consideration towards applicants from disadvantaged backgrounds or schools. Finally, standardized test results can be correlated with background factors, reflecting student environment outside of school. The strategy here is to select a representation of control variables aligned with similar research including @Wiliam2010, @Geiser2007, and @Savolainen2008.

SBAC tests can be administered up to 8 times throughout a student's secondary education. During data exploration, multiple approaches were considered for capturing scores that can be compared to college enrollment:

-   Most recent SBAC scores (last score reported in a school district).

-   Last two SBAC scores to smooth single test variations.

-   A simple aggregation of SBAC scores across all grades in a school district.

-   Weighted scores across all grades emphasizing later grades over earlier.

Additionally, as the last grade where the test is administered is 11th grade, there is a difference between the year a student takes the test, and the year a student enrolls in college. To align, assessment data must be shifted by the number of years between the test being administered and the student being eligible for college enrollment. This research assumes standard high school graduation in 12th grade.

Initial evaluation found no significant difference in performance across the four methods of capturing SBAC scores. The final analysis uses the last grade in which a school district administered the SBAC for a specific year, then offsets the year accordingly. Additionally, data was limited to districts where the last testing took place in 8th, 10th, or 11th grade. This accounts for approximately 90% of school district testing across the state during the years examined.

## Outcome Measures

Based on WSAC goals of 70% post-secondary college enrollment [@WSAC2026Plan], this study focuses on enrollment rates of students in 4-year colleges and universities. Note that student success can be measured in many ways, including enrollment in 2-year colleges or technical schools, enlistment in the military, or regular employment and earning a salary at or above median income. I also recognize that if a student is eligible for and even accepted into college, they may choose an alternative direction in their post-secondary outcomes. Here, we choose 4-year college as an opportunity most likely to lead to higher income and long term economic success and most aligned to WSAC goals.

Enrollment is measured in this data as the percentage of students in a given year that have enrolled in their first year of college. In data, this is represented by the variable `enrollment_all` and is differentiated from enrollment rates by gender defined as `enrollment_female` and `enrollment_male`.

## Methodology

Through regression analysis, I will analyze SBAC test results in respect to college enrollment rates for school districts throughout Washington state. In thsi analysis, multiple forms of regression are considered based on the shape and content of the data. Additionally, I will examine the effect of Math scores on those same enrollment rates to determine if there is a significant difference in influence. I have chosen four different forms of regression to compare opportunities for asessing enrollment rates based on literacy achievement:

1.  Simple linear regression based on **Ordinary Least Squares (OLS)** - provides a baseline for all other analysis.
2.  **Quadratic regression** - used to more accurately match the shape of the data as demonstrated in the Analysis section of this document.
3.  **Fixed Effects regression** to account for hidden variables that may influences differences across school districts and across school years.
4.  **Quadratic regression with control variables** - extending the quadratic regression results to directly control for socioeconomic and demographic variables.

Of these four methods, fixed effects and regression with controls are expected to best account for OVB. As outlined in several previous studies [@Geiser2007; @Hernandez2011], low-income rates, unemployment, and additional factors may confound or otherwise impact SBAC test assessment as captured in `ela_met`. Through a comparison of multiple regression techniques, we can identify a means to track school district success at improving enrollment.

# Analysis

Starting with summary statistics, we can immediately see several interesting patterns emerge in the SBAC assessment data. The tables below examine results for all school districts across all years for which data was collected. In @tbl-assessment we have results for the percent of students meeting expectations on the SBAC ELA test (`ela_met`), along with the SBAC Math test (`math_met`) for comparison. Additionally, data is provided for the rate at which students meet Level 4 standards for both ELA (`ela_lvl4`) and Math (`math_lvl4`). As a reminder, SBAC results are categorized as Level 1 to 4, where levels 3 and 4 indicate students meeting expectations for their grade level.

In @tbl-assessment, the median rate at which students meet expectations for the ELA test (0.63) is more than double the rate for the math test (0.28), indicating a higher rate of reading & writing literacy over math literacy. The Level 4 rates drop off substantially in both cases, as expected, though ELA (0.20) is still more than double the math rate (0.09).

```{r}
#| label: tbl-assessment
#| echo: false
#| tbl-cap: "Summary Statistics - SBAC ELA & Math Assessment"
#| tbl-pos: H

# variables to be summarized
vars_assessment <- c(
  "ela_met", "ela_lvl4", "math_met", "math_lvl4"
)

# Build the summary data set for display in a table
summary_df <- df_assess_final_adj %>%
  summarise(
    across(
      all_of(vars_assessment),
      list(
        median = ~median(.x, na.rm = TRUE),
        min    = ~min(.x, na.rm = TRUE),
        max    = ~max(.x, na.rm = TRUE),
        n      = ~sum(!is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to   = c("variable", "stat"),
    names_pattern = "(.*)_(median|min|max|n)$",
    values_to  = "value"
  ) %>%
  pivot_wider(
    names_from  = stat,
    values_from = value
  ) %>%
  mutate(
    across(c(median, min, max), ~ round(.x, 2)),
    n = as.integer(n),
    variable = factor(variable, levels = vars_assessment)
  ) %>%
  arrange(variable) %>%
  select(variable, n, median, min, max)

# Convert to gt table
summary_df %>%
  gt() %>%
  opt_table_font(size = 10) %>%   # reduce font size
  tab_options(
    table.width = pct(85)   # Scale table to 90% of text width
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_label(
    variable = "Variable",
    n        = "N",
    median   = "Median",
    min      = "Min",
    max      = "Max"
  ) %>%
  tab_style(   # format column headings with bold
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = 1)
  )

```

Focusing on just the ELA assessment rates, and examining them across school years (@tbl-elamet), our range of values across school districts show extensive variance. Median values range from 0.37 in 2014-15 to 0.74 in the following school year, 2015-16. The following three years demonstrate a slow decline from that peak, followed by a sudden drop in the post-pandemic years, from 0.67 in 2018-19 to 0.6 and lower in the 2021-22 and later school years. Note that no testing was conducted during the two COVID pandemic school years.

```{r}
#| label: tbl-elamet
#| echo: false
#| tbl-cap: Summary Statistics - ELA Assessment by School Year
#| tbl-pos: H

# Build summary statistics for ela_met by school year
summary_year <- df_assess_final_adj %>%
  group_by(school_year) %>%
  summarise(
    median = median(ela_met, na.rm = TRUE),
    min    = min(ela_met, na.rm = TRUE),
    max    = max(ela_met, na.rm = TRUE),
    n      = sum(!is.na(ela_met)),
    .groups = "drop"
  ) %>%
  # round numeric summaries
  mutate(
    across(c(median, min, max), ~ round(.x, 2)),
    n = as.integer(n)
  ) %>%
  # reshape so stats are rows and years are columns
  pivot_longer(
    cols = c(median, min, max, n),
    names_to = "stat",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = school_year,
    values_from = value
  ) %>%
  arrange(factor(stat, levels = c("n", "median", "min", "max")))

# Convert to gt table
summary_year %>%
  gt(rowname_col = "stat") %>%
  opt_table_font(size = 10) %>%   # reduce font size
  tab_options(
    table.width = pct(95)   # Scale table to 90% of text width
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_label(
    stat = ""
  ) %>%
  tab_stubhead(label = "School Year") %>%
  tab_style(   # format column headings with bold
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  ) %>%
  # Format the n row with no decimals
  fmt_number(
    columns = everything(),
    rows = stat == "n",
    decimals = 0
  )

```

Interestingly, maximum achievement rates saw less varaince across the school years, never dropping below 0.90 even after the pandemic. The implication is that school districts performing at the top of the range have consistent results, while school districts across the state have varying outcomes from year to year. The boxplots in @fig-boxelamet further emphasize this activity: after a sudden boost in scores in 2015-16, we see a slow decline year over year with a sudden drop in 2021-22..

```{r}
#| label: fig-boxelamet
#| echo: false
#| fig-cap: "Rate of School Districts Meeting SBAC ELA Standards"

# create a temporary data set for the plot
#df_long <- df_assess_final_adj %>%
#  select(school_year, ela_met) %>%
#  pivot_longer(cols = c(enrollment_all, ela_met),
#               names_to = "variable", values_to = "value") %>%
#  mutate(variable = factor(variable, levels = c("enrollment_all", "ela_met")))


# Boxplot
p1 <- ggplot(data = df_assess_final_adj,
             mapping = aes(x = factor(school_year), y = ela_met)) +
  geom_boxplot(color = "darkgray", fill = "darkred") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  labs(
    #title = "Rate of School Districts Meeting SBAC ELA Standards",
    #subtitle = "Last grade tested in school district",
    x = "School Year",
    y = "% Students Meeting Standards"
  )
p1
```

College enrollment demonstrates a mroe stable trend in the tables below. @tbl-enrollment provides an initial breakdown across all school districs and enrollment years considered, with maximum enrollment rates relatively even across both genders. However, median enrolllmnent rates across districts show a clear indication that there are higher enrollment rates for women (`enrollment_female`) than men (`enrollment_male`).

```{r}
#| label: tbl-enrollment
#| echo: false
#| tbl-cap: Summary Statistics - College Enrollment
#| tbl-pos: H

# variables to be summarized
vars_enrollment <- c(
  "enrollment_all", "enrollment_female", "enrollment_male"
)

# Build the summary data set for display in a table
summary_df <- df_enroll_year %>%
  summarise(
    across(
      all_of(vars_enrollment),
      list(
        median = ~median(.x, na.rm = TRUE),
        min    = ~min(.x, na.rm = TRUE),
        max    = ~max(.x, na.rm = TRUE),
        n      = ~sum(!is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to   = c("variable", "stat"),
    names_pattern = "(.*)_(median|min|max|n)$",
    values_to  = "value"
  ) %>%
  pivot_wider(
    names_from  = stat,
    values_from = value
  ) %>%
  mutate(
    across(c(median, min, max), ~ round(.x, 2)),
    n = as.integer(n),
    variable = factor(variable, levels = vars_enrollment)
  ) %>%
  arrange(variable) %>%
  select(variable, n, median, min, max)

# Convert to gt table
summary_df %>%
  gt() %>%
  opt_table_font(size = 10) %>%   # reduce font size
  tab_options(
    table.width = pct(85)   # Scale table to 90% of text width
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_label(
    variable = "Variable",
    n        = "N",
    median   = "Median",
    min      = "Min",
    max      = "Max"
  ) %>%
  tab_style(   # format column headings with bold
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = 1)
  )

```

Gender may have an important role in analyzing enrollment and will be considered in our regression analysis below. Looking at college enrollment by year in @tbl-enrollall, we see consistent results for all four statistics across all enrollment years, especially with median enrollment rates hovering slightly above and below 0.25 or 25% of students. Unfortunately, this is substantially lower than the 70% rate of enrollment targeted by the WSAC.

```{r}
#| label: tbl-enrollall
#| echo: false
#| tbl-cap: Summary Statistics - College Enrollment by Year
#| tbl-pos: H

# Build summary statistics for ela_met by year_grad
summary_year <- df_enroll_year %>%
  group_by(year) %>%
  summarise(
    median = median(enrollment_all, na.rm = TRUE),
    min    = min(enrollment_all, na.rm = TRUE),
    max    = max(enrollment_all, na.rm = TRUE),
    n      = sum(!is.na(enrollment_all)),
    .groups = "drop"
  ) %>%
  # round numeric summaries
  mutate(
    across(c(median, min, max), ~ round(.x, 2)),
    n = as.integer(n)
  ) %>%
  # reshape so stats are rows and years are columns
  pivot_longer(
    cols = c(median, min, max, n),
    names_to = "stat",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = year,
    values_from = value
  ) %>%
  arrange(factor(stat, levels = c("n", "median", "min", "max")))

# Convert to gt table
summary_year %>%
  gt(rowname_col = "stat") %>%
  opt_table_font(size = 10) %>%   # reduce font size
  tab_options(
    table.width = pct(95)   # Scale table to 95% of text width
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_label(
    stat = ""
  ) %>%
  tab_stubhead(label = "Year") %>%
  tab_style(   # format column headings with bold
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_stub()
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_stubhead()
  ) %>%
  # Format the n row with no decimals
  fmt_number(
    columns = everything(),
    rows = stat == "n",
    decimals = 0
  )
```

Interestingly school districts reaching the maximum enrollment rate saw only a small reduction in 2020, the primary pandemic year, to 0.75, down from 0.80 the prior year. Yet 2024 saw the lowest range with maximum rates reaching only 064.

Combining the data requires shifting the SBAC test results by 1, 2, or 4 years to match enrollment. For most school districts, and across most years, the last grade in which SBAC tests were administered occurred in 11th, 10th, or 8th grade. For this analysis, I have identified the last grade for a school district in a year, and shifted that data by the appropriate number of years to match with enrollment. For example, in districts where SBAC was administered in 11th grade for the last time for a cohort, the students in that grade graduated and would enroll in college one year later. This assumes no "gap" years prior to enrollment.

Note that some school districts have years in which the last administration of the SBAC occurred earlier than 8th grade. However, focusing on 8th, 10th, and 11th grades retains approximately 92.5% of the assessment results and captures results more relevant to student achievement closer to graduation and college enrollment. With this adjustment, we have 2,112 district/year combinations for SBAC test assessment across 318 unique school districts in Washington State. Similarly, between 2015 and 2025, we have 2,431 district/year combinations for college enrollment across 229 school districts. When joining this data, based on shifting years to match enrollment, we lose a portion of the assessment results and enrollment results as seen in @tbl-combined.

```{r}
#| label: tbl-combined
#| echo: false
#| tbl-cap: Summary Statistics - Combined College Enrollment and SBAC Data
#| tbl-pos: H

# variables to be summarized
vars_combined <- c(
  "enrollment_all", "ela_met", "ela_lvl4", "math_met", "math_lvl4"
)

# Build the summary data set for display in a table
summary_df <- dfCapLastGradeAdj %>%
  summarise(
    across(
      all_of(vars_combined),
      list(
        median = ~median(.x, na.rm = TRUE),
        min    = ~min(.x, na.rm = TRUE),
        max    = ~max(.x, na.rm = TRUE),
        n      = ~sum(!is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to   = c("variable", "stat"),
    names_pattern = "(.*)_(median|min|max|n)$",
    values_to  = "value"
  ) %>%
  pivot_wider(
    names_from  = stat,
    values_from = value
  ) %>%
  mutate(
    across(c(median, min, max), ~ round(.x, 2)),
    n = as.integer(n),
    variable = factor(variable, levels = vars_combined)
  ) %>%
  arrange(variable) %>%
  select(variable, n, median, min, max)

# Convert to gt table
summary_df %>%
  gt() %>%
  opt_table_font(size = 10) %>%   # reduce font size
  tab_options(
    table.width = pct(85)   # Scale table to 90% of text width
  ) %>%
  cols_align(
    align = "center",
    columns = everything()
  ) %>%
  cols_label(
    variable = "Variable",
    n        = "N",
    median   = "Median",
    min      = "Min",
    max      = "Max"
  ) %>%
  tab_style(   # format column headings with bold
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(   # format first column with bold
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = 1)
  )
```

In @tbl-combined, enrollment variables (`enrollment_all`) dropped to 1,528 records compared to our 2,388 previously examined, while SBAC ELA results (`ela_met`) were reduced to 1,389 records compared to 1,847 before joining the data. This shows a 36% data loss for enrollment data and a 25% data loss for SBAC assessment results due to mismatch in assessment results and enrollment. The total number of records remaining, though, allows sufficient data to complete the analysis.

@fig-scatter is a visualization of the relationship between SBAC ELA assessment results and 4-year college enrollment. Each point is an individual school district for a specific school year. Although there is a strong relationship between higher enrollment rates and higher rates of SBAC ELA achievement, we also see a percentage of school districts at the lower end of SBAC achievement with high enrollment rates, emphasizing a quadratic relationship in the data.

```{r}
#| label: fig-scatter
#| echo: false
#| fig-cap: College Enrollment and ELA Assessment

p1 <- ggplot(data = dfCapLastGradeAdj, 
             aes(x = ela_met, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = NULL,
    x = "% Meeting SBAC ELA Standards",
    y = "% Enrolling in College"
  )
p1
```

As a comparison point, in @fig-lvl4 we can also examine enrollment in relation to the more difficult Level 4 standard (expectations for student achievement are Level 3 or 4).

```{r}
#| label: fig-lvl4
#| echo: false
#| fig-cap: "College Enrollment and ELA Level 4Assessment"

p1 <- ggplot(data = dfCapLastGradeAdj, 
             aes(x = ela_lvl4, y = enrollment_all)) +
  geom_point(alpha = 0.4, color = "darkgray") +
  geom_smooth(method = "auto", se = FALSE,
              linewidth = 1, color = "steelblue") +
  theme_minimal() +
  theme(plot.subtitle = element_text(face="bold")) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    subtitle = NULL,
    x = "% Meeting Level 4 of SBAC ELA",
    y = "% Enrolling in College"
  )
p1
```

The curve here is more complex with a solid line of districts at the left side, not meeting Level 4 standards while still seeing some college enrollment. This implies an unclear relationship between the Level 4 standard and college enrollment which we can explore further with regression.

Previous tables and charts show that variables beyond the SBAC results, e.g. gender, may influence college enrollment, and that the relationship between enrollment and SBAC achievement rates is non-linear. Additionally, there are unknown differences across both school district and school years (or enrollment years) that imply the possible need for entity-fixed and time-fixed effects to avoid OVB. To account for possible variations, Table 6, below includes multiple regressions including linear, quadratic using the R built-in `lm()` function along with `plm()` for panel regression. The fourth regression shown is quadratic with the inclusion of several control variables provided by the SBAC assessment data and US Census Bureau data including gender, employment, median income, and health insurance.

```{r}
#| echo: false

# Linear Regression
model_linear <- lm(
  enrollment_all ~ ela_met,
  data = dfCapLastGradeAdj
)

# Quadratic Regression
model_quad <- lm(
  enrollment_all ~ ela_met + I(ela_met^2),
  data = dfCapLastGradeAdj
)

# Create a pdata.frame with entity and time indexes
pdata_lastgrade <- pdata.frame(dfCapLastGradeAdj,
                               index = c("district_code", "year"))
# Two-way Fixed Effects Regression
model_twoway <- plm(
  enrollment_all ~ ela_met,
  data = pdata_lastgrade,
  model = "within",
  effect = "twoways"
)

# Quadratic Model with control variables
model_quad_controls <- lm(
  enrollment_all ~ ela_met + I(ela_met^2) +
    ela_li_ratio + ela_female_ratio + employment_rate + 
    med_income_household + health_insurance_rate + female_rate,
  data = dfCapLastGradeAdj
)

# Build a table with the output from export_summs
tbl <- export_summs(model_linear, model_quad,
                    model_twoway, model_quad_controls,
                    to = "huxtable",
                    model.names = c("Linear", "Quadratic",
                                    "2-Way Fixed", "Controls"),
                    robust = "HC1",
                    coefs = c(
                      "Intercept" = "(Intercept)",
                      "ELA Met" = "ela_met",
                      "ELA Met^2" = "I(ela_met^2)",
                      "ELA Low Income" = "ela_li_ratio",
                      "ELA Female" = "ela_female_ratio",
                      "Employment (pop.)" = "employment_rate",
                      "Med. Household Income" = "med_income_household",
                      "Health Insurance (pop.)" = "health_insurance_rate",
                      "Female (pop.)" = "female_rate"
                    )
                    )

caption(tbl) <- "Regression of Enrollment by SBAC Assessment"
label(tbl) <- "tbl-regression"
latex_float(tbl) <- "H"

tbl <- set_font_size(tbl, 8)
tbl
```

In the above comparison, the linear and quadratic models demonstrate significance at the 99% level and above across most regressors. Notable exceptions are controls for health insurance rate within the population and gender ratio across the population. Despite gender within the broader district population not showing significance, gender among the rate of those taking the SBAC tests does show significance with a coefficient of 0.04.

Despite the apparent panel shape of the source data, with school districts as entities combined with SBAC assessment and college enrollment tracked across multiple years, 2-way fixed-effects regression shows no significance for ELA assessment results. Additionally, this returned an R-squared value of 0, indicating the model does not explain any variance in the data. Linear regression shows a relatively small R-squared at 0.10, while the quadratic model shows 0.32, matching expectations demonstrated in @fig-scatter. This is boosted by the introduction of control variables in the fourth column, with an R-squared of 0.42. This R-squared result, combined with significance demonstrated by variables for Low Income test takers, employment rates, and median income indicate the quadratic model for ELA assessment rates have a relationship with college enrollment.

As a point of comparison, Table 7 compares the quadratic model results with similar models applied to the Level 4 assessment rates for the SBAC ELA exam, as well as the assessment rates for the SBAC Math exam. Note that control variables determined as not significant have been removed from these models.

```{r}
#| echo: false

# Quadratic Model with control variables
model_quad_controls <- lm(
  enrollment_all ~ ela_met + I(ela_met^2) +
    ela_li_ratio + ela_female_ratio + employment_rate + 
    med_income_household,
  data = dfCapLastGradeAdj
)

# Quadratic Model on ELA Level 4
model_quad_lvl4 <- lm(
  enrollment_all ~ ela_lvl4 + I(ela_lvl4^2) +
    ela_li_ratio + ela_female_ratio + employment_rate + 
    med_income_household,
  data = dfCapLastGradeAdj
)

# Qaudratic model on SBAC Math results
model_quad_math <- lm(
  enrollment_all ~ math_met + I(math_met^2) +
    ela_li_ratio + ela_female_ratio + employment_rate + 
    med_income_household,
  data = dfCapLastGradeAdj
)

# Build a table with the output from export_summs
tbl <- export_summs(model_quad_controls, model_quad_lvl4,
                    model_quad_math,
                    to = "huxtable",
                    model.names = c("ELA SBAC", "ELA Level 4",
                                    "Math SBAC"),
                    robust = "HC1",
                    coefs = c(
                      "Intercept" = "(Intercept)",
                      "ELA Met" = "ela_met",
                      "ELA Met^2" = "I(ela_met^2)",
                      "ELA Low Income" = "ela_li_ratio",
                      "ELA Female" = "ela_female_ratio",
                      "Employment (pop.)" = "employment_rate",
                      "Med. Household Income" = "med_income_household",
                      "ELA Level 4" = "ela_lvl4",
                      "ELA Level 4^2" = "I(ela_lvl4^2)",
                      "Math Met" = "math_met",
                      "Math Met^2" = "I(math_met^2)"
                    )
                    )

caption(tbl) <- "Regression of Enrollment by SBAC Assessment"
label(tbl) <- "tbl-regression"
latex_float(tbl) <- "H"

tbl <- set_font_size(tbl, 8)
tbl

```

All three models in Table 7 produce similar results in terms of significant regressors. The initial model on ELA rate of meeting standards (ela_met) is a slight improvement over the other two models based on R-squared results.

# Discussion and Conclusion

The findings of this article indicate that SBAC ELA assessment rates can be used as a tool to measure expected college enrollment rates for 4-year colleges in Washington State. ELA assessment rates, however, need to be considered in conjunction with a broader range of socioeconomic and demographic variable. The research here indicates income, employment, and gender are also important considerations, supported by @Geiser2007, @Savolainen2008, @Westall2023, and @Hernandez2011.

There are, inevitably, an unlimited range of variables that can influence both the results of standardized tests and the likelihood of enrollment in college. It is not possible for any analysis of this type to consider every control. Key demographics considered by @Talwar2022, @James2024, @Ozek2025 and @Westall2023 include race/ethnicity as well the percentage of students where English is not their first language. Recommended further research on this topic should also include parents' education [@Geiser2007; @Jackson2005; @James2024] and occupation [@Savolainen2008].

As Washington state and the Washington Student Achievement Council work towards execucting against the 2026 Strategic Action Plan [@WSAC2026Plan], knowing where students are at with skill achievement is a critical first step. Just knowing the numbers, however, is not enough. We need to understand how those numbers impact the stated goals. With a defined relationship between assessment and enrollment, WSAC and state legislators can work toward ensuring all districts receive the funding and resources necessary.

Note that I am not attempting to indicate that standardized testing such as the SBAC is the best means to monitor literacy achievement. There may be alternative means of monitoring progress in student literacy capabilities through secondary education. Additionally, there are other standardized tests provided in certain situations. For example, the Washington Access to Instruction & Measurement (WA-AIM) provides testing for students with cognitive disabilities. However, in lieu of a more developed solution, the SBAC standardized tests provide the best option available in Washington schools at this time.

An important gap in the current research is understanding enrollment in two-year colleges and technical schools. This information is available from the enrollment data downloaded and would provide a more complete view of student enrollment after secondary school within the state. Because WSAC goals are specifically related to college enrollment, targeting 70% of graduates, this research specifically excluded students enrolling in the military after high school and students gaining employment at or above median income rates. Both perspectives would provide meaningful information on student outcomes from ELA achievement.

Additionally, this research focused on a point in time for student literacy achievement, specifically the results of SBAC tests taken toward the end of their secondary education. Literacy is a skill set that is established early and encouraged throughout a school experience. Student trends from grade to grade need to be considered to understand if the earliest skills assessments or later assessments are more indicative of future success in a post-secondary experience.

Despite these opportunities for improvement, closer evaluation of SBAC ELA results can provide helpful insight into both the success of literary programs in Washington schools, as well as success in increasing rates of college enrollment across all school districts. To support the entire secondary school population, school administrators need to look beyond traditional college preparatory programs for more opportunities to encourage student enrollment. Establishing a strong base in literacy skills is essential to building that encouragement, and measuring achievement is one of many tools for identifying opportunities.

# References
